---
title: "Reproducible Research: Peer Assessment 1"
author: "Bhanuteja Aryasomayajula"
date: "6/21/2018"
output: html_document
---

```{r}
# Loading the necessary packages
library(knitr)
library(dplyr)
library(ggplot2)
```


```{r}
opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

```{r}
# loading the data

repdata1 <- read.csv('activity.csv')
head(repdata1)
str(repdata1)

#removing the missing values for steps

repdata <- repdata1[!is.na(repdata1$steps),]
head(repdata)
str(repdata)
```

## What is mean total number of steps taken per day?

```{r}
# grouping the data by the date variable
date_wise <- group_by(repdata, date)

#calculating the total number of steps per day
steps_by_day <- summarise(date_wise, total = sum(steps))
steps_by_day

#Plotting a histogram of the total number of steps taken each day
hist(steps_by_day$total, main="Histogram of total number of steps per day", 
     xlab="Total number of steps in a day")
     

#Calculating the mean and median of the steps per day
summary(steps_by_day)   

```
Mean & Median of total number of steps per day are 10766 and 10765 respectively.


## What is the average daily activity pattern?

```{r}
# aggregating the data by interval
steps_by_interval <- aggregate(steps ~ interval, repdata, mean)

# create a time series plot 
plot(steps_by_interval$interval, steps_by_interval$steps, type='l', 
     main="Average number of steps over all days", xlab="Interval", 
     ylab="Average number of steps")
     
# find row with max of steps
max_steps_row <- which.max(steps_by_interval$steps)

# find interval with this max
steps_by_interval[max_steps_row, ]
```
The interval 835 has the maximum average number of steps which is 206.1698


## Imputing missing values

```{r}
# Using the mean for that 5-minute interval where the value is missing for imputing the variable

data_imputed <- repdata1
for (i in 1:nrow(data_imputed)) {
  if (is.na(data_imputed$steps[i])) {
    interval_value <- data_imputed$interval[i]
    steps_value <- steps_by_interval[
      steps_by_interval$interval == interval_value,]
    data_imputed$steps[i] <- steps_value$steps
  }
}
```
The new data set repdata_impute which equals to repdata but without NA’s. All NA’s are replaced with the mean for that 5-minute interval where the value is missing.



# calculate  total number of steps taken each day and plotting the histogram again

```{r}
df_imputed_steps_by_day <- aggregate(steps ~ date, data_imputed, sum)
head(df_imputed_steps_by_day)

hist(df_imputed_steps_by_day$steps, main="Histogram of total number of steps per day (imputed)", 
     xlab="Total number of steps in a day")
     
# get mean and median of imputed data
mean(df_imputed_steps_by_day$steps)
median(df_imputed_steps_by_day$steps)

# get mean and median of data without NA's
mean(steps_by_day$total)
median(steps_by_day$total)

```

Mean values stays the same but therer is slight difference in meadian value.


# Are there differences in activity patterns between weekdays and weekends?

```{r}
data_imputed['type_of_day'] <- weekdays(as.Date(data_imputed$date))
data_imputed$type_of_day[data_imputed$type_of_day  %in% c('Saturday','Sunday') ] <- "weekend"
data_imputed$type_of_day[data_imputed$type_of_day != "weekend"] <- "weekday"

# convert type_of_day from character to factor
data_imputed$type_of_day <- as.factor(data_imputed$type_of_day)

# calculate average steps by interval across all days
df_imputed_steps_by_interval <- aggregate(steps ~ interval + type_of_day, data_imputed, mean)

# creat a plot
qplot(interval, 
      steps, 
      data = df_imputed_steps_by_interval, 
      type = 'l', 
      geom=c("line"),
      xlab = "Interval", 
      ylab = "Number of steps", 
      main = "") +
  facet_wrap(~ type_of_day, ncol = 1)
```

